---
title: "Final"
author: "Your Name"
date: "2025-12-06"
format: html
---

# Introduction

Evictions represent a critical indicator of neighborhood instability and social vulnerability.  
In this project, our **dependent variable** is:

**Quarterly eviction total for 2024, aggregated at the census tract level in Philadelphia.**

Our core research question is:

**Can we predict quarterly eviction filing counts at the census-tract level 1–3 months (i.e., 1 quarter) in advance, enabling the City of Philadelphia to proactively deploy rental assistance, legal aid resources, and community outreach?**

Spatial Unit: Census tract
Temporal Unit: Quarterly
Outcome: Number of eviction filings per tract per quarter
Model: Poisson / Negative Binomial (count outcome), or logistic if we binarize high-risk vs low-risk tracts


**Why Use Quarterly Data?**

Monthly eviction filings are often noisy and volatile.
Aggregating to quarters (3-month periods) smooths short-term fluctuations and better aligns with: Budget cycles, Program planning cycles, Policy reporting (quarterly dashboards)

We therefore build a tract–quarter modeling dataset, while still allowing lag structures that represent 1, 2, 4 quarters in the past.

We incorporate three major categories of predictors:

1. **Structural characteristics** from ACS and property assessment data  
2. **Spatial environmental features** such as crime and transit accessibility  
3. **Temporal dynamics**, including:
   - **Lag 1 quarter** (previous quarter’s evictions)
   - **Lag 2 = Lag 4 quarters** (same quarter in the previous year)

Our goal is to build and compare multiple statistical and spatial models to determine which approach most accurately predicts tract-level eviction activity.


# Data Preparation and Cleaning

This section describes the construction of the final analysis dataset, integrating multiple sources.

```{r}
library(tidycensus)
library(sf)
library(dplyr)
library(readr)
library(zoo)
library(lubridate)
library(ggplot2)
library(stringr)
library(tidyverse)
```

## Census Tracks
```{r}
philly_tracts <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = "B01001_001",  # total population, used just to get geometry
  year = 2022,
  geometry = TRUE
)

philly_tracts <- philly_tracts %>% 
  dplyr::select(GEOID, geometry)
```


## Dependent Variable: 2024 Quarterly Eviction Totals
Eviction filing data from 2021–2024 were collected and aggregated to the tract–quarter level.  
Quarterly lags were computed to represent short-term momentum (lag 1) and annual seasonal effects (lag 2 / lag 4).

```{r}
evictions_raw <- read.csv("data/philadelphia_weekly_2020_2021.csv")

evictions_quarterly <- evictions_raw %>%
  mutate(
    week_date = ymd(week_date),
    year = year (week_date),
    quarter = quarter(week_date),
    year_quarter = paste0(year, "-Q", quarter)
  ) %>%
  # aggregate to quarterly by tract 
  group_by(GEOID, year_quarter, year, quarter, racial_majority, filings_avg,filings_avg_prepandemic_baseline, filings_2020) %>%
  summarize(
    eviction_filings = sum(filings_2020, na.rm = TRUE),
    weeks_in_quarter = n(),
    .groups = "drop"
  )

cat("Quarterly data created:\n")
cat("Rows:", nrow(evictions_quarterly), "\n")
cat("Unique tracts:", n_distinct(evictions_quarterly$GEOID), "\n")
cat("Time periods:", paste(unique(evictions_quarterly$year_quarter), collapse = ", "), "\n\n")
```
## Structural Variables
Structural characteristics include:
- Population, income, poverty indicators, renter share, and other ACS socioeconomic metrics  
- Median property value, low-value housing share, and tax delinquency proxy from OPA assessments  

These variables capture neighborhood housing vulnerability and socioeconomic risk factors.

### ACS factors

```{r}
acs_vars <- c(
  # Demographics
  total_pop          = "B01001_001",
  white_pop          = "B02001_002",
  black_pop          = "B02001_003",
  asian_pop          = "B02001_005",
  hispanic_pop       = "B03003_003",

  # Economic indicators
  median_income      = "B19013_001",
  poverty_total      = "B17001_001",
  poverty_below      = "B17001_002",
  unemployment_total = "B23025_002",
  unemployment_unemployed = "B23025_005",

  # Housing & tenure
  total_housing_units = "B25001_001",
  renter_occupied     = "B25003_003",
  median_rent         = "B25064_001",
  median_home_value   = "B25077_001",

  # Rent burden distribution
  rent_30_to_35_pct   = "B25070_007",
  rent_35_to_40_pct   = "B25070_008",
  rent_40_to_50_pct   = "B25070_009",
  rent_50_plus_pct    = "B25070_010",

  # Housing age & vacancy
  median_year_built   = "B25037_001",
  vacant_units        = "B25002_003",

  # Education - Less than high school (B15003 table)
  edu_total           = "B15003_001",
  less_than_hs_2      = "B15003_002",
  less_than_hs_3      = "B15003_003",
  less_than_hs_4      = "B15003_004",
  less_than_hs_5      = "B15003_005",
  less_than_hs_6      = "B15003_006",
  less_than_hs_7      = "B15003_007",
  less_than_hs_8      = "B15003_008",
  less_than_hs_9      = "B15003_009",
  less_than_hs_10     = "B15003_010",
  less_than_hs_11     = "B15003_011",
  less_than_hs_12     = "B15003_012",
  less_than_hs_13     = "B15003_013",
  less_than_hs_14     = "B15003_014",
  less_than_hs_15     = "B15003_015",
  less_than_hs_16     = "B15003_016"
)

acs_raw <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = acs_vars,
  year = 2022,
  output = "wide",
  geometry = FALSE
)

acs_final <- acs_raw %>%
  mutate(
    # -------------------------
    # Race composition
    # -------------------------
    pct_white     = white_popE     / total_popE * 100,
    pct_black     = black_popE     / total_popE * 100,
    pct_hispanic  = hispanic_popE  / total_popE * 100,
    pct_asian     = asian_popE     / total_popE * 100,

    # -------------------------
    # Poverty & unemployment
    # -------------------------
    poverty_rate      = poverty_belowE / poverty_totalE * 100,
    unemployment_rate = unemployment_unemployedE / unemployment_totalE * 100,

    # -------------------------
    # Tenure & rent burden
    # -------------------------
    pct_renter        = renter_occupiedE / total_housing_unitsE * 100,
    rent_burdened     = rent_30_to_35_pctE + rent_35_to_40_pctE +
                        rent_40_to_50_pctE + rent_50_plus_pctE,
    pct_rent_burdened = rent_burdened / renter_occupiedE * 100,
    pct_severe_burden = rent_50_plus_pctE / renter_occupiedE * 100,

    # -------------------------
    # Education – Less than High School
    # -------------------------
    less_than_hs = less_than_hs_2E + less_than_hs_3E + less_than_hs_4E +
                   less_than_hs_5E + less_than_hs_6E + less_than_hs_7E +
                   less_than_hs_8E + less_than_hs_9E + less_than_hs_10E +
                   less_than_hs_11E + less_than_hs_12E + less_than_hs_13E +
                   less_than_hs_14E + less_than_hs_15E + less_than_hs_16E,

    pct_less_than_hs = less_than_hs / edu_totalE * 100,

    # -------------------------
    # Housing age
    # -------------------------
    property_age  = 2024 - median_year_builtE
  ) %>%
  select(
    GEOID,
    total_popE,
    pct_white, pct_black, pct_hispanic, pct_asian,
    poverty_rate, unemployment_rate,
    median_incomeE, pct_renter,
    pct_rent_burdened, pct_severe_burden,
    pct_less_than_hs,              # <- newly added education variable
    median_rentE, median_home_valueE,
    property_age, vacant_unitsE
  ) %>%
  rename(
    total_pop         = total_popE,
    median_income     = median_incomeE,
    median_rent       = median_rentE,
    median_home_value = median_home_valueE,
    vacant_units      = vacant_unitsE
  )

head(acs_final)
write.csv(acs_final, "acs_final.csv", row.names = FALSE)
```

### Property Data
```{r}
opa <- read_csv("data/opa_properties_public.csv")
```
```{r}
# spatial join
opa_xy <- opa %>%
  filter(!is.na(shape)) %>% 
  filter(str_detect(shape, "POINT\\(")) %>%
  mutate(
    shape_clean = str_remove(shape, "SRID=2272;POINT\\("),
    shape_clean = str_remove(shape_clean, "\\)"),
    x = as.numeric(str_split(shape_clean, " ", simplify = TRUE)[,1]),
    y = as.numeric(str_split(shape_clean, " ", simplify = TRUE)[,2])
  )

opa_sf <- st_as_sf(
  opa_xy,
  coords = c("x", "y"),
  crs = 2272,
  remove = FALSE
)

opa_sf <- st_transform(opa_sf, crs = 4326)

philly_tracts <- philly_tracts %>%
  st_transform(4326)

opa_joined <- st_join(
  opa_sf,
  philly_tracts[, c("GEOID", "geometry")],
  join = st_within
)

opa_clean <- opa_joined %>%
  # have census_tract and market_value
  filter(!is.na(census_tract), !is.na(market_value)) %>%
  mutate(
    # low value dummy
    low_value = market_value < 100000,
    # Tax delinquency proxy
    tax_del = (taxable_building == 0 | taxable_land == 0)
  )


property_summary <- opa_clean %>%
  group_by(GEOID) %>%
  summarise(
    median_market_value = median(market_value, na.rm = TRUE),
    pct_low_value = mean(low_value, na.rm = TRUE) * 100,
    pct_tax_delinquent = mean(tax_del, na.rm = TRUE) * 100
  )

```



## Spatial Variables
Spatial features representing public safety and accessibility include:
- **Crime activity** aggregated via spatial joins to census tracts  
- **Transit accessibility**, measured through density of stations within walking distance  

These variables quantify environmental exposure that may influence eviction risk.
```{r}
tract_centroids <- st_centroid(philly_tracts)

transit_raw <- read_csv("data/Transit_Stops.csv")

transit_sf <- transit_raw %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326)

tract_buffer <- st_buffer(tract_centroids, dist = 800)

station_join <- st_join(tract_buffer, transit_sf)

transit_summary <- station_join %>%
  st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarise(n_stations_800m = n())
```
```{r}
crime <- read_csv("data/incidents_part1_part2.csv")

crime_sf <- crime %>%
  filter(!is.na(lat), !is.na(lng)) %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326)

crime_join <- st_join(
  philly_tracts,
  crime_sf,
  join = st_contains
)

crime_summary <- crime_join %>%
  st_drop_geometry() %>% 
  group_by(GEOID) %>%
  summarise(
    total_crimes = n()
  )
```


```{r}
library(jsonlite)
library(dplyr)

# ACS 2022, Table B01003_001E (Total Population)
url <- "https://api.census.gov/data/2022/acs/acs5?get=B01003_001E,NAME&for=tract:*&in=state:42%20county:101"

philly_pop_raw <- fromJSON(url)

philly_pop <- as.data.frame(philly_pop_raw[-1, ], stringsAsFactors = FALSE)

names(philly_pop) <- c("population", "NAME", "state", "county", "tract")

philly_pop <- philly_pop %>%
  mutate(
    population = as.numeric(population),
    GEOID = paste0(state, county, tract)
  ) %>%
  select(GEOID, population)

crime_summary <- crime_summary %>%
  left_join(philly_pop, by = "GEOID") %>%
  mutate(
    crime_rate = total_crimes / (population / 1000)
  )

```


## Temporal Lags
Two temporal lag measures were added:
- **Lag 1 (t–1 quarter):** captures short-term persistence  
- **Lag 2 (t–4 quarters):** captures seasonal effects across years  

Together these create a panel-like structure enabling spatiotemporal modeling.

All variables were merged into a unified census-tract-level dataset for 2024 modeling and performance evaluation.

```{r}
evictions_with_lags <- evictions_quarterly %>%
  arrange(GEOID, year, quarter) %>%
  group_by(GEOID) %>%
  mutate(
    evictions_lag1Q  = lag(eviction_filings, 1),  # 1 quarter ago (~3 months)
    evictions_lag2Q  = lag(eviction_filings, 2),  # 2 quarters (~6 months)
    evictions_lag4Q  = lag(eviction_filings, 4),  # 4 quarters (~12 months)

    # Rolling means in quarters (e.g., average over last 2 or 4 quarters)
    evictions_avg_2Q = rollmean(eviction_filings, k = 2, fill = NA, align = "right"),
    evictions_avg_4Q = rollmean(eviction_filings, k = 4, fill = NA, align = "right")
  ) %>%
  ungroup()

head(evictions_with_lags)
```
```{r}
evictions_2024_raw <- evictions_with_lags %>%
  filter(year == 2024)

numeric_cols <- evictions_2024_raw %>%
  select(where(is.numeric)) %>%
  names()

evictions_2024_quarterly <- evictions_2024_raw %>%
  group_by(GEOID, year_quarter) %>%
  summarise(
    across(all_of(numeric_cols), sum, na.rm = TRUE),
    .groups = "drop"
  )

```


## Merge All Data
```{r}
final_panel <- evictions_2024_quarterly %>%
  # 1. Join ACS structural data
  left_join(acs_final, by = "GEOID") %>%
  
  # 2. Join Property structural indicators
  left_join(property_summary, by = "GEOID") %>%
  
  # 3. Join spatial transit accessibility
  left_join(transit_summary, by = "GEOID") %>%
  
  # 4. Join crime + crime_rate
  left_join(crime_summary, by = "GEOID") %>%
  
  # 5. Join geometry (for mapping only)
  left_join(philly_tracts, by = "GEOID")

# Print summary
cat("\nMerged dataset created successfully.\n")
cat("Rows:", nrow(final_panel), "\n")
cat("Columns:", ncol(final_panel), "\n")
cat("Unique tracts:", n_distinct(final_panel$GEOID), "\n\n")

# Optional: save final merged table
write_csv(st_drop_geometry(final_panel), "data/final_panel_no_geometry.csv")
st_write(final_panel, "data/final_panel.geojson", delete_dsn = TRUE)

```



# Exploration

This section presents exploratory data analysis, focusing on spatial patterns and distributional characteristics.

## Eviction Data Analysis
- Mapping quarterly eviction totals to visualize spatial clustering  
- Distributional summaries of eviction variations across tracts  
- **Local Indicators of Spatial Association (LISA)** to identify hotspots and coldspots  
- Time-series inspection of lag relationships

## Crime and Environmental Context
- Spatial distribution of crime events and crime rates  
- **K-Nearest Neighbor (KNN)** spatial diagnostics for crime concentrations  
- Comparison with eviction clusters to explore co-location patterns

## Structural and Spatial Features
- Mapping ACS variables, property indicators, and transit accessibility  
- Correlation analysis to detect multicollinearity  
- Exploration of spatial autocorrelation in predictors

This exploratory stage guides expectations about which models may best capture the relationships in eviction dynamics.




# Modeling

Our modeling strategy progresses from simple baselines to increasingly realistic representations of temporal dependence, spatial dependence, and distributional properties of eviction filings.
Because eviction data exhibit strong spatial clustering and temporal persistence, we fit **6 models** and compare their performance and theoretical validity.

```{r}
library(spdep)
library(MASS)      # for Negative Binomial
library(spatialreg)

df <- read_csv("data/final_panel_no_geometry.csv")
df <- df %>% drop_na()


# ---- Outcome variable ----
y_var <- "eviction_filings"

# ---- Temporal lag (Lag 1 quarter) ----
lag_var <- "evictions_lag1Q"

# ---- Structural variables ----
acs_vars <- c(
  "pct_black",
  "median_income", 
  "pct_renter",
  "pct_rent_burdened",
  "median_rent",
  "property_age"
)

# ---- Property variables ----
property_vars <- c(
  "median_market_value",
  "pct_low_value",
  "pct_tax_delinquent"
)

# ---- Spatial variables ----
spatial_vars <- c(
  "n_stations_800m",
  "crime_rate"
)

# Combine predictors into one vector
X_vars <- c(acs_vars, property_vars, spatial_vars)

# Build formula for cross-sectional OLS
formula_ols <- as.formula(
  paste(y_var, "~", paste(X_vars, collapse = " + "))
)

# Build formula for lagged OLS
formula_lag <- as.formula(
  paste(y_var, "~", lag_var, " + ", paste(X_vars, collapse = " + "))
)


```

There is no severe multicollinearity issue in our dataset.

```{r}
library(car)

# VIF examination: ALL OKAY
structural_vars <- c(
  "pct_black", "median_income", "pct_renter", "pct_rent_burdened",
  "median_rent", "property_age", "median_market_value", 
  "pct_low_value", "pct_tax_delinquent", "n_stations_800m", 
  "crime_rate", "vacant_units", "poverty_rate"
)

vif_formula <- as.formula(
  paste("eviction_filings ~", paste(structural_vars, collapse = " + "))
)

vif_ols <- lm(vif_formula, data = df)

vif_results <- vif(vif_ols)
print(vif_results)
```


## Model 1 — Baseline OLS (Cross-sectional)
**Specification:**Evict_2024 ~ ACS + property + spatial   
A simple benchmark without temporal or spatial components.Very low explanatory power (R² = 0.26) and cannot capture either temporal persistence or spatial clustering.

```{r}
model1_ols <- lm(formula_ols, data = df)

summary(model1_ols)

```


## Model 2 — Time-lagged OLS
**Specification:**Evict_2024 ~ Time lag + ACS + property + spatial  
Introduce temporal dependence using the previous quarter’s eviction filings.
**Model fit increases dramatically (R² = 0.91), suggesting eviction filings are highly time-lag-dependent.** But all other variables are insignificant and spatial autocorrelation remains in residuals. It is still biased.

```{r}
model2_lag_ols <- lm(formula_lag, data = df)

summary(model2_lag_ols)
```



## Model 3 — Count Data Models (Poisson)

**Specification:**NB(Evict_2024) ~ Time lag + ACS + property + spatial  
Use Poisson regression to address the count nature of eviction filings. Better suited for non-negative, skewed count outcomes; Improves statistical consistency compared to OLS; But still ignores spatial dependence.

```{r}
model3_poisson <- glm(formula_lag, 
                      data = df, 
                      family = poisson())

summary(model3_poisson)

```

## Model 4 — Spatial Lag Models (SAR / SLX)

Account for strong spatial autocorrelation in both outcome and predictors. All key variables show extremely strong spatial autocorrelation (Moran’s I p ≈ 0), **showing that spatial autocorrelation feature is very significant and need to be consider.**

SAR achieves the best AIC among the traditional models, However, most covariates become statistically insignificant, likely due to multicollinearity induced by spatial spillovers

Overall, it captures spatial processes but loses interpretability and still does not handle count distribution.

```{r}
library(spdep)

df_sf <- df %>%
  left_join(philly_tracts, by = "GEOID") %>%
  st_as_sf()

df_sf <- df_sf %>%
  filter(!st_is_empty(geometry)) %>%
  filter(!is.na(geometry))

nb <- poly2nb(df_sf)
lw <- nb2listw(nb, style = "W")

```


### SLX Model:
**Specification:**Evict_2024 ~ Time lag + X + W * X

Captures spatial spillover in predictors.
```{r}
model4_slx <- lmSLX(
  formula_lag,
  data = df,
  listw = lw
)

summary(model4_slx)
```


### SAR Model:
**Specification:**Evict_2024 = ρ (W * Evict_2024) + βX + ε

Captures spatial dependence directly in the outcome.

```{r}
model4_sar <- lagsarlm(
  formula_lag,
  data = df,
  listw = lw
)

summary(model4_sar)
```

## Model 5 — Posson+spatial lag

Combine strengths of Poisson (for count data) and spatial lag (for spillovers).  
It is more realistic for eviction processes: temporal, spatial, outcome, but residual Moran’s I indicates remaining spatial autocorrelation, meaning the model is still misspecified. Moving in the right direction, but spatial heterogeneity is not fully adjusted.

```{r}
num_vars <- c(
  "evictions_lag1Q", 
  "pct_renter", "crime_rate"
)

for (v in num_vars) {
    df[[v]][!is.finite(df[[v]])] <- median(df[[v]], na.rm = TRUE)
}


df$W_evictions_lag1Q <- lag.listw(lw, df$evictions_lag1Q)
df$W_pct_renter    <- lag.listw(lw, df$pct_renter)
df$W_crime_rate  <- lag.listw(lw, df$crime_rate)

```


```{r}
formula_text <- paste(
  "eviction_filings ~ evictions_lag1Q + W_evictions_lag1Q +",
  "poverty_rate + vacant_units + pct_renter + pct_black + median_income + pct_rent_burdened + median_rent + property_age + median_market_value + pct_low_value + pct_tax_delinquent + n_stations_800m + crime_rate +",
  "W_pct_renter + W_crime_rate"
)

formula_st_poisson <- as.formula(formula_text)

model5 <- glm(
  formula_st_poisson,
  family = poisson(),
  data = df
)

summary(model5)
```


## Model 6 — fixing spatial autocorrelation

**Specification:**

FINAL MODEL. 
Fix remaining spatial autocorrelation by adding **Moran’s Eigenvector (ME) terms** to Model 5.

ME terms successfully remove spatial dependence in residuals. **And AIC decreases further, showing best-performing model overall.**

In this model, coefficients become more stable and interpretable, and satisfies temporal, spatial, and distributional assumptions simultaneously

```{r}
lw_mat <- listw2mat(lw)


meig <- spmoran::meigen(lw_mat)

E <- meig$sf[, 1:5]
colnames(E) <- paste0("eig", 1:5)


df <- cbind(df, E)

eig_cols <- paste0("eig", 1:5)

formula_text <- paste(
  "eviction_filings ~ evictions_lag1Q + W_evictions_lag1Q +",
  "poverty_rate + vacant_units + pct_renter + pct_black + median_income + pct_rent_burdened + median_rent + property_age + median_market_value + pct_low_value + pct_tax_delinquent + n_stations_800m + crime_rate +",
  "W_pct_renter + W_crime_rate +",
  paste(eig_cols, collapse = " + ")
)

formula_st_poisson <- as.formula(formula_text)

model6 <- glm(
  formula_st_poisson,
  family = poisson(),
  data = df
)

summary(model6)


```
Model 6 provides the most accurate and theoretically justified explanation of eviction filings because it integrates **temporal dependence (lag); spatial spillovers (W·terms); count modeling (Poisson) and unobserved spatial heterogeneity (eigenvectors).**

This combination corrects model misspecification, reduces AIC, and yields interpretable and robust estimates.



# Model Diagnostics and Validation

To evaluate model performance and generalizability, we conduct:

## Cross-validation on 2024 Quarters
- Train models on earlier quarters  
- Validate predictions on later quarters (e.g., Q3 → Q4)  
- Evaluate RMSE, MAE, R², and residual spatial autocorrelation  

```{r}
library(purrr)

# --- Performance function  ---
perf <- function(actual, predicted) {
  list(
    MAE  = mean(abs(actual - predicted)),
    RMSE = sqrt(mean((actual - predicted)^2)),
    R2   = 1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
  )
}

quarters <- c("2024-Q1", "2024-Q2", "2024-Q3", "2024-Q4")

model_list  <- list(model1_ols, model2_lag_ols, model3_poisson,
                    model5, model6)

model_names <- c("Model1_OLS", "Model2_TimeLag_OLS", "Model3_Poisson",
                 "Model5_ST_Poisson", "Model6_Final")

cv_results_df <- map2_dfr(
  model_list, model_names,
  
  ~{
    mdl <- .x
    mdl_name <- .y
 
    map2_dfr(
      quarters[-1],           # test quarter
      quarters[-length(quarters)],
      ~{
        train_q <- quarters[quarters < .x]
        
        df_train <- df[df$year_quarter %in% train_q, ]
        df_test  <- df[df$year_quarter == .x, ]
  
        preds <- if (mdl_name == "Model4_SAR") {
          predict(lagsarlm(formula(mdl), data = df_train, listw = lw),
                  newdata = df_test, type = "response")
        } else if (inherits(mdl, "glm")) {
          predict(mdl, newdata = df_test, type = "response")
        } else if (inherits(mdl, "lm")) {
          predict(mdl, newdata = df_test)
        }
        
        good <- !is.na(preds)
        
        tibble(
          Quarter = .x,
          MAE  = mean(abs(df_test$eviction_filings[good] - preds[good])),
          RMSE = sqrt(mean((df_test$eviction_filings[good] - preds[good])^2)),
          R2   = 1 - sum((df_test$eviction_filings[good] - preds[good])^2) /
                   sum((df_test$eviction_filings[good] - mean(df_test$eviction_filings[good]))^2),
          model = mdl_name
        )
      }
    )
  }
)

cv_results_df

```



```{r}
write.csv(cv_results_df, "cv_results_df.csv", row.names = FALSE)
write.csv(df, "df.csv", row.names = FALSE)
```


Cross-validation performance across quarters shows a clear improvement as temporal, spatial, and distributional components are progressively added to the models. 

Model 1 performs the worst, confirming that a simple OLS cannot capture eviction dynamics. **Model 2 achieves the best predictive accuracy overall, reflecting the strong temporal persistence in eviction filings; however, because it captures only time-lag effects and leaves severe spatial autocorrelation unresolved, it is not an adequate structural model.**

Poisson-based models (Models 3 and 5) perform more consistently across time by aligning with the count nature of eviction data, and Model 5 further improves by incorporating spatial spillovers. The final Model 6 produces similarly strong predictive performance while also fixing residual spatial dependence through Moran’s eigenvectors, yielding the most theoretically complete and empirically reliable model.

Therefore, despite Model 2’s raw accuracy, Model 6 is the chosen final model because it provides both solid predictive ability and the correct spatiotemporal data-generating structure.


```{r}
cv_plot <- cv_results_df %>%
  pivot_longer(cols = c(MAE, RMSE, R2),
               names_to = "Metric", values_to = "Value")

ggplot(cv_plot, aes(x = model, y = Value, fill = model)) +
  geom_col() +
  facet_grid(Metric ~ Quarter, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Cross-Validation Performance by Quarter and Model",
    x = "Model",
    y = "Value"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face="bold")
  )

```


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

cv_summary <- cv_results_df %>%
  group_by(model) %>%
  summarise(
    MAE  = mean(MAE),
    RMSE = mean(RMSE),
    R2   = mean(R2)
  ) %>%
  pivot_longer(cols = c(MAE, RMSE, R2),
               names_to = "Metric", values_to = "Value")

ggplot(cv_summary, aes(x = model, y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Cross-Validation Performance Comparison (Model 1–6)",
    x = "Model",
    y = "Metric Value",
    fill = "Performance Metric"
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(face="bold")
  )
```



## Residual Diagnostics
- Spatial autocorrelation in residuals (Moran’s I)  
- Heteroskedasticity checks  

```{r}
df$resid <- residuals(model6, type = "pearson")

# Moran's I
moran_test <- moran.test(df$resid, lw)
moran_test

```
```{r}
library(spdep)

lag_resid <- lag.listw(lw, df$resid)

plot_df <- data.frame(
  resid = df$resid,
  lag_resid = lag_resid
)

ggplot(plot_df, aes(x = resid, y = lag_resid)) +
  geom_point(alpha = 0.5, size = 1.8, color = "#2C7BB6") +
  geom_smooth(method = "lm", se = FALSE, color = "#D7191C", linewidth = 1) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Residuals",
    y = "Spatial Lag of Residuals",
    title = "Moran Scatterplot of Model Residuals"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 16)
  )

```

```{r}
library(ggplot2)

df_sf$resid <- residuals(model6, type = "pearson")

ggplot(df_sf) +
  geom_sf(aes(fill = resid), color = NA) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, name = "Residuals"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    panel.grid = element_blank()
  ) +
  labs(title = "Spatial Distribution of Model Residuals")

```
```{r}
library(lmtest)
library(sandwich)

bp_test <- bptest(model6)
bp_test

```
```{r}
df$fit  <- fitted(model6)
df$resid <- residuals(model6, type = "pearson")

ggplot(df, aes(x = fit, y = resid)) +
  geom_point(alpha = 0.5, size = 1.8, color = "#2C7BB6") +  
  geom_smooth(method = "lm", se = FALSE, 
              color = "#D7191C", linewidth = 1) + 
  labs(
    x = "Fitted Values",
    y = "Pearson Residuals",
    title = "Heteroskedasticity Check: Residuals vs Fitted"
  ) +
  theme_minimal(base_size = 14)
```



# Conclusion








